---
title: "Defining Study Populations and Variables"
subtitle: "Quan Summit 2026 — 30 minute workshop"
author: "Elliot Martin"
date: 2026-09-28
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
editor: visual
execute:
  warning: false
  message: false
---

# Quan Summit 2026-09-28: Defining Study Populations and Variables

In this workshop we'll do a mini study looking at the effects of different comorbidities on Heart Failure (HF) readmission. This will involve,

1.  Selecting a HF cohort from the Discharge Abstract Database (DAD)
2.  Determining which patients were readmitted (our outcome)
3.  Identifying the comorbidities of interest in our cohort (our exposures), using the DAD and physician claims
4.  Running a simple survival analysis

The github repo for this code to run in RStudio can be found at: https://github.com/centre-for-health-informatics/QuanSummit2026-DefiningPopulations-rstudio

![RStudio QR Code](assets/QuanSummit2026-rstudio-qr.svg){width="250" fig-align="center"}

A web version of this can be found at https://centre-for-health-informatics.github.io/QuanSummit2026-DefiningPopulations-webr/

![webr QR Code](assets/QuanSummit2026-webr-qr.svg){width="250" fig-align="center"}

## Load/Install necessary modules

```{r}
#| label: load libraries 
  #install.packages(c("readr", "dplyr", "tidyr", "stringr",
  #                  "tibble", "ggplot2", "survival"))

library(readr)     # read_csv, column type control
library(dplyr)     # filter / mutate / group_by / joins
library(tidyr)     # pivot_longer, pivot_wider
library(stringr)   # str_detect for code matching
library(tibble)    # tibble(), tribble()
library(ggplot2)   # plots
library(survival)  # Surv, coxph, survfit, concordance

```

## Load the data and take a quick look

::: callout-important
**All the data used in this notebook has been synthetically generated** - The proportions of different conditions have been engineered for this work and are not based on real data - Claude was used for the synthetic generation script, which works for this illustration, but you shouldn't take away anything from the data generated - The script used to generate the data is `generate-data.R` and can be found in the `code` folder of the git repo - you can use the script to generate additional data sets, or do things like tweak the proportion of HF admissions etc. - the script is expecting the directory structure of the repo, i.e., be run from the `code` folder and save the data to the `data` folder
:::

### Load DAD data

```{r}
#| label: load data

dad <- read_csv(
  "data/dad_synthetic.csv",
  col_types  = cols(
    .default   = col_character(),
    AGE        = col_integer(),
    ADMIT_DTTM = col_datetime("%Y-%m-%d %H:%M"),
    DISCH_DTTM = col_datetime("%Y-%m-%d %H:%M")
  )
)
glimpse(dad[, 1:10])
```

Look at some statistics of the DAD data

```{r}
#| label: Examine DAD
dad |>
  summarise(
    admissions   = n(),
    patients    = n_distinct(PATIENT_ID),
    first_admit = min(ADMIT_DTTM),
    last_admit  = max(ADMIT_DTTM)
  )

```

Look at how many patients had multiple admissions

```{r}
#| label: Admission Distribution
dad |>
  count(PATIENT_ID, name = "admissions") |>       # one row per patient
  count(admissions, name = "patients")
```

### Load Claims data

```{r}
claims <- read_csv(
  "data/claims_synthetic.csv",
  col_types = cols(
    .default     = col_character(),
    SERVICE_DATE = col_date("%Y-%m-%d")
  )
)

glimpse(claims)
```

Look at some statistics for the claims data

```{r}
#| label: Examine Claims
claims |>
  summarise(
    claims        = n(),
    patients      = n_distinct(PATIENT_ID),
    first_service = min(SERVICE_DATE),
    last_service  = max(SERVICE_DATE)
  )
```

## Build the final cohort

### Select all Heart Failure admissions

::: callout-important
The DAD does not use decimal places in ICD-10-CA codes, so I50.0 will be represented as I500
:::

```{r}
# Quan et al. (2005) ICD-10 congestive heart failure codes. 
hf_pattern <- "^I099|^I110|^I130|^I132|^I255|^I420|^I42[5-9]|^I43|^I50|^P290"

# DXCODE1 is the most responsible diagnosis -- the reason for the stay.
hf_admissions <- dad |>
  filter(str_detect(DXCODE1, hf_pattern), AGE >= 18) |>
  select(PATIENT_ID, ENCOUNTER_ID, ADMIT_DTTM, DISCH_DTTM)

hf_admissions |>
  summarise(admissions = n(), patients = n_distinct(PATIENT_ID))
```

### Construct episodes of care

A patient may have multiple entries in the DAD, but they could be considered one "Episode of Care", e.g., a patient moved from a rural site to a larger hospital generates **two DAD entries** but should be considered **one** episode of care.

We are using a simple algorithm here, and considering any admission within 12 hours of discharge to be the same episode of care. CIHI uses a slightly different algorithm, with varying time windows depending on if the transfer was coded or not.

```{r}
#| label: outcome parameters
EPISODE_GAP_HOURS <- 12    # admission < this many hours after discharge = same episode
DEATH_DISP        <- c("07", "72", "73", "74")  # died: pre-2018 code + 2018+ codes 
```

```{r}
#| label: episodes of care
admissions <- dad |>
  filter(PATIENT_ID %in% hf_admissions$PATIENT_ID) |>
  select(PATIENT_ID, ENCOUNTER_ID, ADMIT_CATEGORY,
         ADMIT_DTTM, DISCH_DTTM, DISCH_DISP) |>
  arrange(PATIENT_ID, ADMIT_DTTM) |>
  mutate(
    same_patient     = PATIENT_ID == lag(PATIENT_ID, default = ""),
    hours_since_prev = if_else(same_patient,
                               as.numeric(difftime(ADMIT_DTTM, lag(DISCH_DTTM),
                                                   units = "hours")),
                               NA_real_),
    # Continues the previous episode if admitted within the gap.
    continues_prev = coalesce(hours_since_prev < EPISODE_GAP_HOURS, FALSE),
    episode        = cumsum(!continues_prev)   # a unique id per episode
  )

# An episode STARTS on a row that doesn't continue the previous one, and
# ENDS on a row whose next row doesn't continue it.
episode_start <- admissions |>
  filter(!continues_prev) |>
  select(PATIENT_ID, episode,
         ep_admit = ADMIT_DTTM, ep_category = ADMIT_CATEGORY)  # how it STARTED

episode_end <- admissions |>
  filter(!lead(continues_prev, default = FALSE)) |>
  select(episode, ep_disch = DISCH_DTTM, ep_disp = DISCH_DISP) # how it ENDED

episodes <- episode_start |>
  inner_join(episode_end, by = "episode") |>
  inner_join(admissions |> count(episode, name = "n_admissions"), by = "episode")

tibble(admissions = nrow(admissions), episodes = nrow(episodes))
episodes |> count(n_admissions, name = "episodes")
```

### First HF Admission (Index episode)

```{r}
#| label: index-episode
index <- hf_admissions |>
  inner_join(admissions |> select(ENCOUNTER_ID, episode), by = "ENCOUNTER_ID") |>
  arrange(PATIENT_ID, ADMIT_DTTM) |>
  distinct(PATIENT_ID, .keep_all = TRUE) |>   # keeps each patient's first row
  inner_join(episodes, by = c("PATIENT_ID", "episode")) |> 
  filter(!ep_disp %in% DEATH_DISP)               #excluding dead patients since they can't return
    
glimpse(index)
```

### Determine the outcome for each member of the cohort

This will be whether/when they are readmitted after their index hospitalization.

```{r}
#| label: readmissions
# Readmission = the first UNPLANNED (urgent) episode starting after the index
# episode's discharge. Episodes, not abstracts: a transfer is not a readmit.
readmits <- index |>
  select(PATIENT_ID, ep_disch) |>
  inner_join(
    episodes |> select(PATIENT_ID, r_admit = ep_admit, r_category = ep_category),
    by = "PATIENT_ID"
  ) |>
  filter(r_admit > ep_disch)

# How many later episodes were planned (elective)? These are NOT events.
readmits |> count(r_category)

first_readmit <- readmits |>
  filter(r_category == "U") |>
  arrange(PATIENT_ID, r_admit) |>
  distinct(PATIENT_ID, .keep_all = TRUE) |>   # earliest unplanned episode
  select(PATIENT_ID, first_readmit = r_admit)
```

### Build the final cohort

```{r}
#| label: build-cohort

STUDY_END <- max(dad$ADMIT_DTTM)

cohort <- index |>
  left_join(first_readmit, by = "PATIENT_ID") |>
  mutate(
    days_to_readmit = as.numeric(difftime(first_readmit, ep_disch, units = "days")),
    days_to_end     = as.numeric(difftime(STUDY_END,     ep_disch, units = "days")),
    # Time at risk = min(time to readmission, time to administrative end)
    time_days = pmin(coalesce(days_to_readmit, Inf), days_to_end),
    event     = as.integer(!is.na(days_to_readmit) & days_to_readmit <= time_days)
  ) |>
  filter(time_days > 0)     # remove events with no observable follow-up


cohort |>
  summarise(patients = n(), readmitted = sum(event),
            rate = mean(event), median_days_at_risk = median(time_days))
```

## Determine the exposures for each patient

These will be the comorbidities that we want to check if they affect the outcome. Let's pick a few.

### Age

This will be the easiest variable to get, which we can get from the DAD. It can be computed from the date of admission and birthdate, but unless there is a clear need birthdate will likely not be released due to its identifiable nature. Instead the analyst will generally compute the age for you and release that.

```{r}
#| label: age
# AGE is age at THAT admission. Check: does it change within patient?
dim(cohort)
cohort <- cohort |>
  inner_join(dad |> select(AGE, ENCOUNTER_ID), by = "ENCOUNTER_ID")

dim(cohort)
```

```{r}
#| label: age-plot
ggplot(cohort, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "grey30") +
  labs(title = "Age at index HF admission", x = "Age (years)", y = "Patients") +
  theme_minimal()
```

### Sex

Sex is a categorical variable, but there are a few things to keep in mind.

1.  The categories are not just M/F
2.  The data doesn't do a very good job at disentangling gender and sex

```{r}
#| label: sex

cohort <- cohort |>
  inner_join(dad |> select(SEX, ENCOUNTER_ID), by = "ENCOUNTER_ID") |>
  mutate(SEX = factor(SEX, levels = c("F", "M", "O", "U")))

cohort |> count(SEX, .drop = FALSE)
```

### COPD

This is our first comorbidity which we will look for in the DAD using the presented algorithms.

```{r}
#| label: copd
copd_codes <- "^I27[89]|^J4[0-4]|^J4[67]|^J6[0-7]|^J684|^J70[13]"

cohort <- cohort |>
  inner_join(dad |> 
                filter(ENCOUNTER_ID %in% cohort$ENCOUNTER_ID) |>
                mutate(copd = as.integer(
                 if_any(starts_with("DXCODE"), ~ str_detect(coalesce(.x, ""), copd_codes))
               )) |>
                select(ENCOUNTER_ID, copd),
               by="ENCOUNTER_ID"
               )

mean(cohort$copd)
```

### CKD

This is our second comorbidity which we will look for in the DAD using the presented algorithms.

```{r}
#| label: ckd
ckd_codes <- "^N0[0-9]|^N1[0-9]|^N2[0-3]"

cohort <- cohort |>
  inner_join(dad |> 
                filter(ENCOUNTER_ID %in% cohort$ENCOUNTER_ID) |>
                mutate(ckd = as.integer(
                 if_any(starts_with("DXCODE"), ~ str_detect(coalesce(.x, ""), ckd_codes))
               )) |>
                select(ENCOUNTER_ID, ckd),
               by="ENCOUNTER_ID"
               )

mean(cohort$ckd)
```

### Hypertension

This is the last comorbidity we will look at. As hypertension is not usually diagnosed in hospital, and is generally not directly responsible for a large use of resources in the hospital, it is known to have low sensitivity in the DAD. Therefore for this case we will also use claims data.

```{r}
#| label: htn-linkage

htn_icd10 <- "^I1[0-35]"
htn_icd9 <- "^40[1-5]"

window_days <- 365*2


dad_htn <- cohort |>
  inner_join(dad |> 
                filter(ENCOUNTER_ID %in% cohort$ENCOUNTER_ID) |>
                mutate(htn_dad = as.integer(
                 if_any(starts_with("DXCODE"), ~ str_detect(coalesce(.x, ""), htn_icd10))
               )) |>
                select(ENCOUNTER_ID, htn_dad),
               by="ENCOUNTER_ID"
               ) |>
  select(PATIENT_ID, htn_dad)

claims_htn <- claims |>
  filter(PATIENT_ID %in% cohort$PATIENT_ID) |>
  inner_join(cohort |> select(PATIENT_ID, index_admit = ADMIT_DTTM),
             by = "PATIENT_ID") |>
  filter(SERVICE_DATE < as.Date(index_admit)) |>
  mutate(htn = as.integer(
                 if_any(starts_with("DXCODE"), ~ str_detect(coalesce(.x, ""), htn_icd9))
               )) |>
  arrange(PATIENT_ID, htn, SERVICE_DATE) |>
  mutate(
    k_row     = row_number() + 2 - 1,        # row of the k-th claim
    same_run  = PATIENT_ID[k_row] == PATIENT_ID,    # same patient
    span_days = as.numeric(SERVICE_DATE[k_row] - SERVICE_DATE)
  ) |>
  filter(same_run, span_days <= window_days) |>       # NA past the end drops out
  distinct(PATIENT_ID, htn) |>
  group_by(PATIENT_ID) |>
  summarise(htn_claims = as.integer(any(htn==1)))
  
htn_admin <- dad_htn|>
  inner_join(claims_htn, by = "PATIENT_ID") |>
  mutate(htn = as.integer(htn_dad==1 |htn_claims==1))


cohort <- cohort |> 
  inner_join(htn_admin |> select(PATIENT_ID, htn), by = "PATIENT_ID")

mean(cohort$htn)
```

## Analyze the data

### Let's take a look at our dataframe first and get some statistics we'll need for table 1

```{r}
#| label: table1
# Suppress any count between 1 and 4 at the point the table is built, not
# afterwards by hand.
n_pct <- function(x) {
  n <- sum(x)
  if (n > 0 && n < 5) "<5" else sprintf("%d (%.0f%%)", n, 100 * mean(x))
}

table1 <- cohort |>
  group_by(readmitted = if_else(event == 1L, "Readmitted", "Not readmitted")) |>
  summarise(
    n           = as.character(n()),
    age_mean_sd = sprintf("%.1f (%.1f)", mean(AGE), sd(AGE)),
    female      = n_pct(SEX == "F"),
    sex_O_or_U  = n_pct(SEX %in% c("O", "U")),
    copd        = n_pct(copd == 1),
    ckd         = n_pct(ckd == 1),
    htn         = n_pct(htn == 1)
  )

# Transpose so variables are rows -- the usual Table 1 layout.
table1 |>
  pivot_longer(-readmitted, names_to = "variable") |>
  pivot_wider(names_from = readmitted, values_from = value)
```

### Survival Analysis

#### Split the data into derivation and validation sets

```{r}
#| label: split
set.seed(1234)

# Stratify on the outcome so both halves get the same event rate.
cohort <- cohort |>
  group_by(event) |>
  mutate(split = if_else(row_number() %in% sample(n(), round(0.7 * n())),
                         "derivation", "validation")) |>
  ungroup()

cohort |> count(split, event)
```

#### Fit a model to the derivation set

```{r}
#| label: cox
derivation <- cohort |> filter(split == "derivation") |> droplevels()

fit <- coxph(
  Surv(time_days, event) ~ AGE + SEX + copd + ckd + htn,
  data = derivation
)

summary(fit)
```

```{r}
#| label: hr-table
# Tidy HR table without {broom} (keeps the webR dependency list short).
ci <- summary(fit)$conf.int
cf <- summary(fit)$coefficients

tibble(
  term  = rownames(ci),
  HR    = round(ci[, "exp(coef)"], 2),
  lower = round(ci[, "lower .95"], 2),
  upper = round(ci[, "upper .95"], 2),
  p     = signif(cf[, "Pr(>|z|)"], 2)
)
```

#### Check the model performance on the validation set

```{r}
#| label: validate
validation <- cohort |> filter(split == "validation")

# Harrell's C on data the model has never seen...
concordance(fit, newdata = validation)

# ...versus apparent (over-optimistic) performance on the derivation set.
concordance(fit)
```
